---
  title: "mids-203-project"
output: html_document
date: "2026-04-09"
---
```{r}
# Load libraries
library(ggplot2)

# Read CSV (adjust path as needed)
df <- read.csv("merged_raw_all_years.csv", na.strings = c("nan", "NaN"))
dim(df)
```

```{r}
training_year <- 2006
```

Defining a function to get columns from that year:
```{r}
get_cols <- function(vars,df,year) {
  cols <- colnames(df)[
    grepl(paste(vars, collapse = "|"), colnames(df)) &
      grepl(paste0(year, "$"), colnames(df))
  ]
  return(cols)
}
```

```{r}
# create per capita columns for GDP and emissions

pop_name <- paste0("POP.raw.", training_year)
vars_to_pc <- c("GDP", "CDO", "SO2", "BLC", "NOT")  

for (v in vars_to_pc) {
  raw_name <- paste0(v, ".raw.", training_year)
  pc_name  <- paste0(v, "_PC.raw.", training_year)
  df[[pc_name]] <- df[[raw_name]] / df[[pop_name]]
  print(pc_name)
}



vars_y <- c("PFL","FCL","FSS","RLI","SHI")
#vars_x <- c("CDO", "SO2","BLC", "NOT","GDP")

vars_x <- c(
  "CDO_PC", "SO2_PC", "BLC_PC", "NOT_PC","GDP_PC")

c <- get_cols(vars_x,df,training_year)

plot_histograms <- function(vars,df,title) {
  par(mfrow = c(length(vars), 2), mar = c(2, 2, 2, 1),oma = c(0, 0, 3, 0))

  for (v in vars) {
    var_name <- paste0(v, ".raw.", training_year)
    x <- df[[var_name]]
  
    hist(x,
         main = paste("Raw:", v),
         xlab = var_name)
  
    hist(log(pmax(x,1e-8)), #add a tiny constant
         main = paste("Log:", v),
         xlab = paste0("log(", var_name, ")"))
  }
  mtext(title, outer = TRUE, cex = 1.5, font = 2)
}

plot_histograms(vars_x, df, title = "Raw vs. Log x-variables")
plot_histograms(vars_y, df, title = "Raw vs. Log y-variables")

```
Now I look for collinearity between the variables:

```{r}
# Scatterplot matrix of x-variables and y-variables

# modifying vars_y to only include FSS, FCL, PFL
vars_y <- c("FSS","FCL")

x <- get_cols(vars_x,df,training_year)
y <- get_cols(vars_y,df,training_year)

for (col in c(x,y)) {
  log_name <- paste0("log_",col)
  df[[log_name]] <- log10(pmax(df[[col]], 1e-8))
}

x_log <- paste0("log_", x)
y_log <- paste0("log_", y)

summary(df[x_log])

# Scatterplot matrix of x
x_log_clean <- na.omit(df[, c('iso','code','country', x_log)])
pairs(x_log_clean[x_log])

# Scatterplot matrix of y
y_log_clean <- na.omit(df[, c('iso','code','country', y_log)])
pairs(y_log_clean[y_log])

# x variables
r2_x <- cor(x_log_clean[, x_log])^2
r2_x

# y variables
r2_y <- cor(y_log_clean[, y_log])^2
r2_y
```
```{r}
pca_x <- prcomp(x_log_clean[,x_log], center = TRUE, scale. = TRUE)

var_explained <- pca_x$sdev^2
prop_var_explained <- var_explained / sum(var_explained)
print(prop_var_explained)

lims <- max(abs(pca_x$rotation[,1:2]))

plot(pca_x$rotation[,1], pca_x$rotation[,2],
     xlim = c(-lims, lims),
     ylim = c(-lims, lims),
     xlab = "PC1", ylab = "PC2",
     main = "PCA Loadings Plot",
     pch = 19)

abline(h = 0, v = 0, col = "gray")

text(pca_x$rotation[,1], pca_x$rotation[,2],
     labels = rownames(pca_x$rotation),
     pos = 3, cex = 0.8)

loadings_df <- as.data.frame(pca_x$rotation)
round(loadings_df[, 1:2], 3)  # PC1 and PC2

par(mar = c(8, 4, 4, 2))  # increase bottom margin for labels

barplot(pca_x$rotation[,1],
        main = "PC1 Loadings",
        las = 2,              # makes labels vertical (closest base option)
        cex.names = 0.8)      # shrink labels if needed

barplot(pca_x$rotation[,2],
        main = "PC2 Loadings",
        las = 2,              # makes labels vertical (closest base option)
        cex.names = 0.8)      # shrink labels if needed

```

Conclusions from PCA are that we can reduce dimensionality (esp PC1).

```{r}
# add pc1 and pc2 to log x-variable df
x_log_clean$PC1 <- pca_x$x[,1]
x_log_clean$PC2 <- pca_x$x[,2]
x_log_clean$PC3 <- pca_x$x[,3]
x_log_clean$PC4 <- pca_x$x[,4]
x_log_clean$PC5 <- pca_x$x[,5]

# merge back into y-variable df
df_pc <- merge(df, x_log_clean[, c('iso','code','country','PC1','PC2','PC3','PC4','PC5')], by = c('iso','code','country'))
```

Next, we plot the individual output variables vs. PC1:

```{r}
par(mfrow = c(2, 2))

for (y_var in y) {
  plot(df_pc[[y_var]] ~ df_pc$PC1,
       main = y_var,
       xlab = "PC1",
       ylab = y_var,
       pch = 19)
}



```

Next, fit linear models of output variables vs. PC1:

``` {r}
library(lmtest)

par(mfrow = c(2, 2))  # 1 row, 3 columns

results <- data.frame()

for (y_var in y) {
  
  model <- lm(as.formula(paste(y_var, "~ PC1")), data = df_pc)
  model_sum <- summary(model)
  bp <- bptest(model)
  fstat <- model_sum$fstatistic
  results <- rbind(
      results,
      data.frame(
        y_var = y_var,
        adj_r2 = model_sum$adj.r.squared,
        f_pvalue = pf(fstat["value"], fstat["numdf"], fstat["dendf"], lower.tail = FALSE),
        bp_stat = unname(bp$statistic),
        bp_pvalue = bp$p.value,
        stringsAsFactors = FALSE
      )
    )
  
  }
  
  print(results)

```
Variations on simplest model Y vs. PC1:

``` {r}
par(mfrow = c(2, 2))  # 1 row, 3 columns

for (y_var in y) {
  
  model <- lm(as.formula(paste(y_var, "~ PC1 + PC2 + PC3 + PC4 + PC5")), data = df_log_pc)
  cat("\n", y_var, "\n")
  cat("R²:", summary(model)$r.squared, "\n")
  
  # Print coefficient table (includes p-values)
  print(summary(model))
  
  plot(model, which = 1, main = paste("Residuals vs Fitted:", y_var))
  print(bptest(model))
  
}
```
```{r}
df_log_pc[c(123,91,141,152),'country']


```

Appendix - trying out literally everything via PCA

```{r}
# PCA function

run_pca <- function(x, y, df, rhs = "~ PC1") {
  library(lmtest)
  # Keep needed columns only
  df_log <- df[, c("iso", "code", "country", x, y)]
  df_log[, c(x, y)] <- log(df[, c(x, y)] + 1)
  
  # Optional GDP filter if present
  if ("gdp" %in% names(df_log)) {
    cutoff <- quantile(df_log$gdp, 0.05, na.rm = TRUE)
    df_log <- df_log[df_log$gdp > cutoff, ]
  }
  
  # Remove rows with missing x values before PCA
  x_log_clean <- na.omit(df_log[, c("iso", "code", "country", x)])
  
  # Run PCA
  pca_x <- prcomp(x_log_clean[, x], center = TRUE, scale. = TRUE)
  n_pcs <- min(5, ncol(pca_x$x))
  pc_cols <- paste0("PC", seq_len(n_pcs))
  
  # Add PC scores to x_log_clean
  x_log_clean[, pc_cols] <- pca_x$x[, seq_len(n_pcs), drop = FALSE]
  
  # Merge PCs back into full log dataframe
  df_log_pc <- merge(
    df_log,
    x_log_clean[, c("iso", "code", "country", pc_cols)],
    by = c("iso", "code", "country")
  )
  
  #return(df_log_pc)
  
  # Fit models and make residual plots
  resid_plots <- list()
  models <- list()
  stats_table <- data.frame()
  
  for (y_var in y) {
    model <- lm(as.formula(paste(y_var, rhs)), data = df_log_pc)
    models[[y_var]] <- model
    
    model_summary <- summary(model)
    fstat <- model_summary$fstatistic
    f_pval <- pf(fstat[1], fstat[2], fstat[3], lower.tail = FALSE)
    bp <- bptest(model)
    
    cat("\n", y_var, "\n")
    cat("R²:", model_summary$r.squared, "\n")
    cat(
      "F-statistic:",
      round(fstat[1], 3),
      "on", fstat[2], "and", fstat[3], "DF",
      "p-value:", format(f_pval, scientific = TRUE, digits = 3), "\n"
    )
    print(bp)
    
    plot_df <- data.frame(
      fitted = fitted(model),
      residuals = resid(model)
    )
    
    resid_plots[[y_var]] <- ggplot(plot_df, aes(x = fitted, y = residuals)) +
      geom_point() +
      geom_hline(yintercept = 0, linetype = "dashed") +
      labs(
        title = paste("Residuals vs Fitted:", y_var),
        x = "Fitted values",
        y = "Residuals"
      ) +
      theme_minimal()
  }
  
  wrap_plots(resid_plots, ncol = 2)
  
}
```

```{r}

eco_biodiversity <- c(
  "MKP",  # Marine KBA Protection
  "MHP",  # Marine Habitat Protection
  "MPE",  # Marine Protection Stringency
  "PAR",  # Protected Areas Representativeness Index
  "SPI",  # Species Protection Index
  "TBN",  # Terrestrial Biome Protection
  "TKP",  # Terrestrial KBA Protection
  "PAE",  # Protected Area Effectiveness
  "PHL",  # Protected Human Land
  "RLI",  # Red List Index
  "SHI",  # Species Habitat Index
  "BER"   # Bioclimatic Ecosystem Resilience
)

eco_forests <- c(
  "PFL",  # Primary Forest Loss
  #"IFL",  # Intact Forest Landscape Loss
  "FCL",  # Tree cover loss weighted by permanency
  "TCG",  # Net change in tree cover
  "FLI"   # Forest Landscape Integrity
)

eco_fisheries <- c(
  "FSS",  # Fish Stock Status
  "FCD",  # Fish Catch Discarded
  "BTZ",  # Bottom Trawling in EEZ
  "BTO",  # Bottom Trawling in Global Ocean
  "RMS"   # Regional Marine Trophic Index
)

eco_air_pollution <- c(
  "OEB",  # Ozone exposure in KBAs
  "OEC",  # Ozone exposure in croplands
  "NXA",  # Adjusted NOx growth rate
  "SDA"   # Adjusted SO2 growth rate
)

eco_agriculture <- c(
  "SNM",  # Sustainable Nitrogen Management
  "PSU",  # Phosphorus Surplus
  "PRS",  # Pesticide Pollution Risk
  "RCY"   # Relative Crop Yield
)

eco_water <- c(
  "WWG",  # Wastewater generated
  "WWC",  # Wastewater collected
  "WWT",  # Wastewater treated
  "WWR"   # Wastewater reused
)

eco_all <- c(
  eco_biodiversity,
  eco_forests,
  eco_fisheries,
  eco_air_pollution,
  eco_agriculture,
  eco_water
)

eco_minus_y <- c(setdiff(eco_all, vars_y))

env_health_air <- c(
  "HPE",  # Anthropogenic PM2.5 exposure
  "HFD",  # Household solid fuels
  "OZD",  # Ozone exposure
  "NOD",  # NOx exposure
  "SOE",  # SO2 exposure
  "COE",  # CO exposure
  "VOE"   # VOC exposure
)
```

```{r}
vars_x_2 <- c(climate_raw)
vars_y_2 <- c("PFL","FCL","FSS")

x <- get_cols(vars_x_2, df, training_year)
y <- get_cols(vars_y_2, df, training_year)

run_pca(x,y,df,rhs = "~ PC1")
run_pca(x,y,df,rhs = "~ PC1 + PC2 + PC3")
```

```{r}
summary(df[, get_cols(eco_minus_y, df, training_year)
])
```

----------- THINGS THAT DIDN'T WORK --------------------------------------------------------------------

Let's see if the poor fit is because we're not grouping the countries very well. After all, landlocked countries won't have much result for fisheries, and coastal countries might have less active deforestation as it won't be a main industry.

```{r}
country_list <- df_log_pc$country

landlocked <- c(
  "Afghanistan", "Armenia", "Azerbaijan", "Bhutan", "Bolivia",
  "Botswana", "Burkina Faso", "Burundi", "Central African Republic",
  "Chad", "Czech Republic", "Ethiopia", "Hungary", "Kazakhstan",
  "Kyrgyzstan", "Laos", "Lesotho", "Luxembourg", "Malawi",
  "Mali", "Moldova", "Mongolia", "Nepal", "Niger",
  "North Macedonia", "Paraguay", "Rwanda", "Serbia",
  "Slovakia", "South Sudan", "Eswatini", "Switzerland",
  "Tajikistan", "Turkmenistan", "Uganda", "Uzbekistan",
  "Zambia", "Zimbabwe", "Austria", "Belarus"
)

coastal <- setdiff(country_list, landlocked)

df_log_pc$type <- ifelse(
  df_log_pc$country %in% landlocked, "landlocked",
  ifelse(df_log_pc$country %in% coastal, "coastal", NA)
)

for (y_var in y) {
  p <- ggplot(df_log_pc, aes(x = PC1, y = .data[[y_var]], color = type)) +
    geom_point() +
    facet_wrap(~ type) 
    labs(
      title = y_var,
      x = "PC1",
      y = y_var,
      color = "Type"
    ) +
    theme_minimal()
  print(p)
}
```
Not seeming like there's much improvement to the model from separating out countries based on a crude categorization. It also doesn't seem like FSS contains any data from landlocked countries (probably why it's missing so many rows).

What about continents?

```{r}
library(countrycode)

df_log_pc$continent <- countrycode(
  df_log_pc$iso,
  origin = "iso3c",
  destination = "continent"
)

# now plot faceted by continent

for (y_var in y) {
  p <- ggplot(df_log_pc, aes(x = PC1, y = .data[[y_var]])) +
    geom_point() +
    facet_wrap(~ continent) 
    labs(
      title = y_var,
      x = "PC1",
      y = y_var,
      color = "Continent"
    ) +
    theme_minimal()
  print(p)
}
```
Maturity

```{r}
df$POP.raw.1960
```

----- ARCHIVED: DATA EXPLORATION -------------------------

- Realizing that a lot of the variables in the dataset are missing like 50% of the rows...so let's first rank all the possible Y variables (in 'ecosystem vitality on page 3') and X variables (in 'climate change on page 4') based on how many NA rows there are, then choose based on a tradeoff between relevance and # of missing rows.

```{r}
ecosystem_vars <- c(
  # Biodiversity & Habitat
  "MKP", "MHP", "MPE", "PAR", "SPI", "TBN", "TKP", "PAE", "PHL", "RLI", "SHI", "BER",
  
  # Forests
  "PFL", "IFL", "FCL", "TCG", "FLI",
  
  # Fisheries
  "FSS", "FCD", "BTZ", "BTO", "RMS",
  
  # Air Pollution
  "OEB", "OEC", "NXA", "SDA",
  
  # Agriculture
  "SNM", "PSU", "PRS", "RCY",
  
  # Water Resources
  "WWG", "WWC", "WWT", "WWR"
)

eco_col_matches <- colnames(df)[
  grepl(paste(ecosystem_vars, collapse = "|"), colnames(df)) &
    grepl(paste0(training_year,"$"), colnames(df))
]

df_ecosystem = df[, eco_col_matches]

# sort each column in the ecosystem vitality (ECO) category by the number of NAs in the year 2019
sort(colSums(is.na(df_ecosystem)))

```
Most relevant Y-variables would be:
- PFL (primary forest loss) - 137 missing rows out of 220
    - Alternative: FCL (74 missing rows)
- FSS (fish stock status) - 102 missing rows out of 220
    - Alternative: RLI/SHI - 2 and 43 missing rows

Now to look at the x-variables:
```{r}
climate_indicators <- c(
  # Indicators in Climate Change Mitigation
  "CDA",  # Adjusted emissions growth rate for CO2
  "CDF",  # CO2 growth rate (country-specific targets)
  "CHA",  # Adjusted emissions growth rate for methane
  #"FGA",  # Adjusted emissions growth rate for F-gases
  "NDA",  # Adjusted emissions growth rate for nitrous oxide
  "BCA",  # Adjusted emissions growth rate for black carbon
  "LUF",  # Net carbon fluxes due to land cover change
  "GTI",  # GHG growth rate adjusted by emissions intensity
  "GTP",  # GHG growth rate adjusted by per capita emissions
  "GHN",  # Projected emissions in 2050
  "CBP"   # Projected cumulative emissions to 2050 relative to carbon budget
)

climate_raw <- c(
  # Raw variables for climate change
  "CDO", # CO2 emissions [Gg], excluding land use and forestry
  "CH4", # Methane emissions [Gg]
  "FOG", # F-gasses emissions [Gg CO2-eq.]
  "NOT", # N2O emissions [Gg]
  "SO2", # SO2 emissions [Gg]
  "BLC", # Black carbon emissions [Gg]
  "LUE", # CO2 net fluxes (emissions and sinks) from land use and land cover change
  "GDP"
  )

get_na <- function(df, vars, label, year = training_year) {
  cols <- colnames(df)[
  grepl(paste(vars, collapse = "|"), colnames(df)) &
    grepl(paste0(year, "$"), colnames(df))
  ]
  df_subset = df[, cols]
  cat(label, ":\n")
  print(sort(colSums(is.na(df_subset))))
  cat("\n")
}

get_na(df, climate_indicators, "Number of NA in climate indicator variables:")
get_na(df, climate_raw, "Number of NA in climate raw variables:")
```
Use these raw variables:
- CDO (carbon dioxide emissions) - related to fossil fuel combustion, primary output of industrial energy use
- SO2 - industrial (coal, combustion)
- BLC - fossil fuel combustion (soot)
- NOT (N2O emissions) - mostly comes from agriculture (fertilizer) but some from industrial/combustion
- GDP

These can all be converted to per capita by dividing by POP (population)

Do not use:
- FOG (F-gas emissions) - industrial, but missing a lot of data
- CH4 (methane emissions) - less industry, more agriculture
- LUE (land-use emissions) - more about land transformation, often for agriculture

Next step is to run a PCA on the x-variables and y-variables to try to reduce the number of features:
- X-variables: CDO/SO2/BLC/NOT/GDP
- Y-variables: PFL/FCL/FSS/RLI/SHI

First, it's time to plot the variables and see if they need some transformation first:
```{r}
summary(df$FSS.raw.2012)
```

