---
  title: "mids-203-project"
output: html_document
date: "2026-04-09"
---
  ```{r cars}
# Load libraries
library(ggplot2)

# Read CSV (adjust path as needed)
df <- read.csv("merged_raw_all_years.csv")
dim(df)
```

First task: examine the data for cleanliness issues

- Realizing that a lot of the variables in the dataset are missing like 50% of the rows...so let's first rank all the possible Y variables (in 'ecosystem vitality on page 3') and X variables (in 'climate change on page 4') based on how many NA rows there are, then choose based on a tradeoff between relevance and # of missing rows.
- I'm using 2019, which is the last year with fisheries data (see page 130/131)

```{r}
ecosystem_vars <- c(
  # Biodiversity & Habitat
  "MKP", "MHP", "MPE", "PAR", "SPI", "TBN", "TKP", "PAE", "PHL", "RLI", "SHI", "BER",
  
  # Forests
  "PFL", "IFL", "FCL", "TCG", "FLI",
  
  # Fisheries
  "FSS", "FCD", "BTZ", "BTO", "RMS",
  
  # Air Pollution
  "OEB", "OEC", "NXA", "SDA",
  
  # Agriculture
  "SNM", "PSU", "PRS", "RCY",
  
  # Water Resources
  "WWG", "WWC", "WWT", "WWR"
)

eco_col_matches <- colnames(df)[
  grepl(paste(ecosystem_vars, collapse = "|"), colnames(df)) &
    grepl("2019$", colnames(df))
]

df_ecosystem = df[, eco_col_matches]

# sort each column in the ecosystem vitality (ECO) category by the number of NAs in the year 2019
sort(colSums(is.na(df_ecosystem)))

```
Most relevant Y-variables would be:
- PFL (primary forest loss) - 137 missing rows out of 220
    - Alternative: FCL (74 missing rows)
- FSS (fish stock status) - 102 missing rows out of 220
    - Alternative: RLI/SHI - 2 and 43 missing rows

Now to look at the x-variables:
```{r}
climate_indicators <- c(
  # Indicators in Climate Change Mitigation
  "CDA",  # Adjusted emissions growth rate for CO2
  "CDF",  # CO2 growth rate (country-specific targets)
  "CHA",  # Adjusted emissions growth rate for methane
  "FGA",  # Adjusted emissions growth rate for F-gases
  "NDA",  # Adjusted emissions growth rate for nitrous oxide
  "BCA",  # Adjusted emissions growth rate for black carbon
  "LUF",  # Net carbon fluxes due to land cover change
  "GTI",  # GHG growth rate adjusted by emissions intensity
  "GTP",  # GHG growth rate adjusted by per capita emissions
  "GHN",  # Projected emissions in 2050
  "CBP"   # Projected cumulative emissions to 2050 relative to carbon budget
)

climate_raw <- c(
  # Raw variables for climate change
  "CDO", # CO2 emissions [Gg], excluding land use and forestry
  "CH4", # Methane emissions [Gg]
  "FOG", # F-gasses emissions [Gg CO2-eq.]
  "NOT", # N2O emissions [Gg]
  "SO2", # SO2 emissions [Gg]
  "BLC", # Black carbon emissions [Gg]
  "LUE" #CO2 net fluxes (emissions and sinks) from land use and land cover change
  )

get_na <- function(df, vars, label, year = "2019") {
  cols <- colnames(df)[
  grepl(paste(vars, collapse = "|"), colnames(df)) &
    grepl(paste0(year, "$"), colnames(df))
  ]
  df_subset = df[, cols]
  cat(label, ":\n")
  print(sort(colSums(is.na(df_subset))))
  cat("\n")
}

get_na(df, climate_indicators, "Number of NA in climate indicator variables:")
get_na(df, climate_raw, "Number of NA in climate raw variables:")

# Also look at how much data is missing for GDP

cat("Number of NA in raw GDP data:", sum(is.na(df[, "GDP.raw.2019"])), "\n")

```
Use these raw variables:
- CDO (carbon dioxide emissions) - related to fossil fuel combustion, primary output of industrial energy use
- SO2 - industrial (coal, combustion)
- BLC - fossil fuel combustion (soot)
- NOT (N2O emissions) - mostly comes from agriculture (fertilizer) but some from industrial/combustion
- GDP

Do not use:
- FOG (F-gas emissions) - industrial, but missing a lot of data
- CH4 (methane emissions) - less industry, more agriculture
- LUE (land-use emissions) - more about land transformation, often for agriculture

Let's now see what combinations of best X and best Y's have the most data

```{r}

vars_y <- c("PFL","FCL","FSS","RLI","SHI")
vars_x <- c("CDO", "SO2","BLC", "NOT","GDP")

get_clean_dim <- function(df, y_var, x_vars = vars_x, year = "2019") {
  cols <- colnames(df)[
    grepl(paste(c(y_var, x_vars), collapse = "|"), colnames(df)) &
      grepl(paste0(year, "$"), colnames(df))
  ]
  
  df_subset <- df[, c("code", "iso", "country", cols), drop = FALSE]
  df_clean <- na.omit(df_subset)

  return(dim(df_clean))
}

# Loop through y variables
for (var in vars_y) {
  dims <- get_clean_dim(df, var)
  cat("Y =", var, ":", dims[1], "rows", "\n")
}

```
- For forest: see if PFL and FCL correlate well; if so, use FCL
- For biodiversity: see if FSS/RLI/SHI correlate well; if not, FSS still has a lot of data

Next step is to run a PCA on the x-variables and y-variables to try to reduce the number of features:
- X-variables: CDO/SO2/BLC/NOT/GDP
- Y-variables: PFL/FCL/FSS/RLI/SHI

First, it's time to plot the variables and see if they need some transformation first:

```{r}

plot_histograms <- function(vars,title) {
  par(mfrow = c(length(vars), 2), mar = c(2, 2, 2, 1),oma = c(0, 0, 3, 0))

  for (v in vars) {
    var_name <- paste0(v, ".raw.2019")
    x <- df[[var_name]]
  
    hist(x,
         main = paste("Raw:", v),
         xlab = var_name)
  
    hist(log10(x[x > 0]),
         main = paste("Log10:", v),
         xlab = paste0("log10(", var_name, ")"))
  }
  mtext(title, outer = TRUE, cex = 1.5, font = 2)
}

plot_histograms(vars_x, "Raw vs. Log10 x-variables")
plot_histograms(vars_y, "Raw vs. Log10 y-variables")

```


Run a PCA on the x-variables:

```{r}
climate_vars <- c(
  # Climate Change Mitigation
  "CDA",  # Adjusted emissions growth rate for CO2
  "CDF",  # CO2 growth rate (country-specific targets)
  "CHA",  # Adjusted emissions growth rate for methane
  "FGA",  # Adjusted emissions growth rate for F-gases
  "NDA",  # Adjusted emissions growth rate for nitrous oxide
  "BCA",  # Adjusted emissions growth rate for black carbon
  "LUF",  # Net carbon fluxes due to land cover change
  "GTI",  # GHG growth rate adjusted by emissions intensity
  "GTP",  # GHG growth rate adjusted by per capita emissions
  "GHN",  # Projected emissions in 2050
  "CBP"   # Projected cumulative emissions to 2050 relative to carbon budget
)

raw_vars <- c("CDO", "CH4", "FOG", "NOT", "BLC", "GDP", "POP", "LUE")

all_x <- c(climate_vars, raw_vars)

climate_col_matches <- colnames(df)[
  grepl(paste(all_x, collapse = "|"), colnames(df)) &
    grepl("2019$", colnames(df))
]

df_climate = df[, climate_col_matches]
df_climate_clean <- na.omit(df_climate)

pca_climate <- prcomp(df_climate_clean, center = TRUE, scale. = TRUE)

plot(pca_climate$rotation[,1], pca_climate$rotation[,2],
     xlab = "PC1", ylab = "PC2",
     main = "PCA Loadings Plot",
     pch = 19)

text(pca_climate$rotation[,1], pca_climate$rotation[,2],
     labels = rownames(pca_climate$rotation),
     pos = 3, cex = 0.8)

loadings_df <- as.data.frame(pca_climate$rotation)
round(loadings_df[, 1:2], 3)  # PC1 and PC2

par(mar = c(8, 4, 4, 2))  # increase bottom margin for labels

barplot(pca_climate$rotation[,1],
        main = "PC1 Loadings",
        las = 2,              # makes labels vertical (closest base option)
        cex.names = 0.8)      # shrink labels if needed

barplot(pca_climate$rotation[,2],
        main = "PC2 Loadings",
        las = 2,              # makes labels vertical (closest base option)
        cex.names = 0.8)      # shrink labels if needed

```


```{r}
plot(log(GDP.raw.2019) ~ log(CDO.raw.2019),
     data = df,
     xlab = "log(CO2 Emissions)",
     ylab = "log(GDP)",
     main = "Log-Log GDP vs CO2 (2019)",
     pch = 19)

plot(log(FSS.raw.2019) ~ log(CDO.raw.2019),
     data = df,
     xlab = "log(CO2 Emissions)",
     ylab = "log(FSS)",
     main = "Log-Log FSS vs CO2 (2019)",
     pch = 19)

plot(log(PFL.raw.2019) ~ log(CDO.raw.2019),
     data = df,
     xlab = "log(CO2 Emissions)",
     ylab = "log(PFL)",
     main = "Log-Log PFL vs CO2 (2019)",
     pch = 19)
```

plot some distributions

```{r}
hist(log(df$CDO.raw.2019))
hist(df$FSS.raw.2019)
hist(df$PFL.raw.2019)
```


```{r}
plot(log(PFL.raw.2019) ~ CDA.raw.2019,
     data = df,
     xlab = "log(CDA)",
     ylab = "log(PFL)",
     main = "Log-Log PFL vs CDA (2019)",
     pch = 19)
```

