---
title: "mids-203-project"
output: html_document
date: "2026-04-09"
---

```{r cars}
# Load libraries
library(ggplot2)

# Read CSV (adjust path as needed)
df <- read.csv("merged_raw_all_years.csv")
dim(df)
```

Added column names bc i got a little confused -Malachi
```{r}
colnames(df)[grepl("PFL|FSS|CDA|CHA|FGA|NDA", colnames(df))]
```
First task: examine the data for cleanliness issues

- Realizing that a lot of the variables in the dataset are missing like 50% of the rows...so let's first rank all the possible Y variables (in 'ecosystem vitality on page 3') and X variables (in 'climate change on page 4') based on how many NA rows there are, then choose based on a tradeoff between relevance and # of missing rows.
- I'm using 2019, which is the last year with fisheries data (see page 130/131)

```{r}
ecosystem_vars <- c(
  # Biodiversity & Habitat
  "MKP", "MHP", "MPE", "PAR", "SPI", "TBN", "TKP", "PAE", "PHL", "RLI", "SHI", "BER",
  
  # Forests
  "PFL", "IFL", "FCL", "TCG", "FLI",
  
  # Fisheries
  "FSS", "FCD", "BTZ", "BTO", "RMS",
  
  # Air Pollution
  "OEB", "OEC", "NXA", "SDA",
  
  # Agriculture
  "SNM", "PSU", "PRS", "RCY",
  
  # Water Resources
  "WWG", "WWC", "WWT", "WWR"
)

eco_col_matches <- colnames(df)[
  grepl(paste(ecosystem_vars, collapse = "|"), colnames(df)) &
    grepl("2019$", colnames(df))
]

df_ecosystem = df[, eco_col_matches]

# sort each column in the ecosystem vitality (ECO) category by the number of NAs in the year 2019
sort(colSums(is.na(df_ecosystem)))

```
Most relevant Y-variables would be:
  - PFL (primary forest loss) - 137 missing rows out of 220
- Alternative: FCL (74 missing rows)
- FSS (fish stock status) - 102 missing rows out of 220
- Alternative: RLI/SHI - 2 and 43 missing rows
- Could consider using TBN (terrestrial biome protection/biodiversity/ecosystem vitality) - not ideal but it has no missing data

Now to look at the x-variables:
  
```{r}
climate_vars <- c(
  # Climate Change Mitigation
  "CDA",  # Adjusted emissions growth rate for CO2
  "CDF",  # CO2 growth rate (country-specific targets)
  "CHA",  # Adjusted emissions growth rate for methane
  "FGA",  # Adjusted emissions growth rate for F-gases
  "NDA",  # Adjusted emissions growth rate for nitrous oxide
  "BCA",  # Adjusted emissions growth rate for black carbon
  "LUF",  # Net carbon fluxes due to land cover change
  "GTI",  # GHG growth rate adjusted by emissions intensity
  "GTP",  # GHG growth rate adjusted by per capita emissions
  "GHN",  # Projected emissions in 2050
  "CBP"   # Projected cumulative emissions to 2050 relative to carbon budget,
)

climate_col_matches <- colnames(df)[
  grepl(paste(climate_vars, collapse = "|"), colnames(df)) &
    grepl("2019$", colnames(df))
]

df_climate = df[, climate_col_matches]

# sort each column in the ecosystem vitality (ECO) category by the number of NAs in the year 2019
sort(colSums(is.na(df_climate)))


```

The x-variables we might use are basically all missing 20-30 rows, not bad. We can then choose based purely on what's most relevant:
- CDO (carbon dioxide emissions)
- CH4 (methane emissions)
- FOG (F-gas emissions)
- NOT (N2O emissions)

We can also investigate how much missing data is in the GDP for 2019:
```{r}
sum(is.na(df[, 'GDP.raw.2019']))
```

What if we looked at only our best variables of interest and saw how small that dataset would be?

```{r}
vars_y <- c("PFL", "FSS")
vars_x <- c("CDO", "CH4", "FOG", "NOT")
vars_z <- c("GDP")
patterns <- c(vars_y,vars_x,vars_z)


cols_matched <- colnames(df)[
  grepl(paste(patterns, collapse = "|"), colnames(df)) &
  grepl("2019$", colnames(df))
]

df_2019 <- df[, c("code", "iso", "country", cols_matched)]
# How much data are we left with if we remove all rows with NA in any of these columns of interest?

df_2019_clean <- na.omit(df_2019)
dim(df_2019_clean)
summary(df_2019_clean[, cols_matched])
```
It appears that for 2019, which is the last year that we still have fisheries data (FSS), we only have 43 countries that have data for all the variables of interest.

Checking pairwise for countries if we just do a direct X and Y relationship
```{r}
vars_y <- c("PFL", "FSS")
vars_x <- c("CDA", "CHA", "FGA", "NDA")
control <- "GDP"
year <- "2019"

pair_2019 <- do.call(rbind, lapply(vars_y, function(yv) {
  do.call(rbind, lapply(vars_x, function(xv) {
    cols <- c(
      paste0(yv, ".raw.", year),
      paste0(xv, ".raw.", year),
      paste0(control, ".raw.", year)
    )
    
    temp <- df[, c("country", cols)]
    
    data.frame(
      y_var = yv,
      x_var = xv,
      year = as.integer(year),
      n_countries = nrow(na.omit(temp))
    )
  }))
}))

pair_2019[order(-pair_2019$n_countries), ]
```

Checking Year overlaps
```{r}
vars_y <- c("PFL", "FSS")
vars_x <- c("CDA", "CHA", "FGA", "NDA")

get_years <- function(var, data) {
  cols <- colnames(data)[grepl(paste0("^", var, "\\.raw\\."), colnames(data))]
  sub(paste0("^", var, "\\.raw\\.(\\d{4})$"), "\\1", cols)
}

count_pair_nonmissing <- function(y_var, x_var, year, data) {
  y_col <- paste0(y_var, ".raw.", year)
  x_col <- paste0(x_var, ".raw.", year)
  
  if (!(y_col %in% colnames(data)) || !(x_col %in% colnames(data))) return(NA_integer_)
  
  sum(!is.na(data[[y_col]]) & !is.na(data[[x_col]]))
}

pair_results <- do.call(rbind, lapply(vars_y, function(yv) {
  do.call(rbind, lapply(vars_x, function(xv) {
    
    overlap_years <- sort(intersect(get_years(yv, df), get_years(xv, df)))
    
    if (length(overlap_years) == 0) {
      return(data.frame(
        y_var = yv,
        x_var = xv,
        n_overlap_years = 0,
        overlap_years = NA_character_,
        best_year = NA_character_,
        best_n_complete = NA_integer_
      ))
    }
    
    counts <- sapply(overlap_years, function(yr) count_pair_nonmissing(yv, xv, yr, df))
    best_idx <- which.max(counts)
    
    data.frame(
      y_var = yv,
      x_var = xv,
      n_overlap_years = length(overlap_years),
      overlap_years = paste(overlap_years, collapse = ", "),
      best_year = overlap_years[best_idx],
      best_n_complete = counts[best_idx]
    )
  }))
}))

pair_results[order(-pair_results$best_n_complete), ]
```
