---
title: "mids-203-project"
output: html_document
date: "2026-04-09"
---

```{r cars}
# Load libraries
library(ggplot2)
library(car)
library(lmtest)

# Read CSV (adjust path as needed)
df <- read.csv("merged_raw_all_years.csv")
dim(df)
```

Added column names bc i got a little confused -Malachi
```{r}
colnames(df)[grepl("PFL|FSS|CDA|CHA|FGA|NDA", colnames(df))]
```
First task: examine the data for cleanliness issues

- Realizing that a lot of the variables in the dataset are missing like 50% of the rows...so let's first rank all the possible Y variables (in 'ecosystem vitality on page 3') and X variables (in 'climate change on page 4') based on how many NA rows there are, then choose based on a tradeoff between relevance and # of missing rows.
- I'm using 2019, which is the last year with fisheries data (see page 130/131)

```{r}
ecosystem_vars <- c(
  # Biodiversity & Habitat
  "MKP", "MHP", "MPE", "PAR", "SPI", "TBN", "TKP", "PAE", "PHL", "RLI", "SHI", "BER",
  
  # Forests
  "PFL", "IFL", "FCL", "TCG", "FLI",
  
  # Fisheries
  "FSS", "FCD", "BTZ", "BTO", "RMS",
  
  # Air Pollution
  "OEB", "OEC", "NXA", "SDA",
  
  # Agriculture
  "SNM", "PSU", "PRS", "RCY",
  
  # Water Resources
  "WWG", "WWC", "WWT", "WWR"
)

eco_col_matches <- colnames(df)[
  grepl(paste(ecosystem_vars, collapse = "|"), colnames(df)) &
    grepl("2019$", colnames(df))
]

df_ecosystem = df[, eco_col_matches]

# sort each column in the ecosystem vitality (ECO) category by the number of NAs in the year 2019
sort(colSums(is.na(df_ecosystem)))

```
Most relevant Y-variables would be:
  - PFL (primary forest loss) - 137 missing rows out of 220
- Alternative: FCL (74 missing rows)
- FSS (fish stock status) - 102 missing rows out of 220
- Alternative: RLI/SHI - 2 and 43 missing rows
- Could consider using TBN (terrestrial biome protection/biodiversity/ecosystem vitality) - not ideal but it has no missing data

Now to look at the x-variables:
  
```{r}
climate_vars <- c(
  # Climate Change Mitigation
  "CDA",  # Adjusted emissions growth rate for CO2
  "CDF",  # CO2 growth rate (country-specific targets)
  "CHA",  # Adjusted emissions growth rate for methane
  "FGA",  # Adjusted emissions growth rate for F-gases
  "NDA",  # Adjusted emissions growth rate for nitrous oxide
  "BCA",  # Adjusted emissions growth rate for black carbon
  "LUF",  # Net carbon fluxes due to land cover change
  "GTI",  # GHG growth rate adjusted by emissions intensity
  "GTP",  # GHG growth rate adjusted by per capita emissions
  "GHN",  # Projected emissions in 2050
  "CBP"   # Projected cumulative emissions to 2050 relative to carbon budget,
)

climate_col_matches <- colnames(df)[
  grepl(paste(climate_vars, collapse = "|"), colnames(df)) &
    grepl("2019$", colnames(df))
]

df_climate = df[, climate_col_matches]

# sort each column in the ecosystem vitality (ECO) category by the number of NAs in the year 2019
sort(colSums(is.na(df_climate)))


```

The x-variables we might use are basically all missing 20-30 rows, not bad. We can then choose based purely on what's most relevant:
- CDO (carbon dioxide emissions)
- CH4 (methane emissions)
- FOG (F-gas emissions)
- NOT (N2O emissions)

We can also investigate how much missing data is in the GDP for 2019:
```{r}
sum(is.na(df[, 'GDP.raw.2019']))
```

What if we looked at only our best variables of interest and saw how small that dataset would be?

```{r}
vars_y <- c("PFL", "FSS")
vars_x <- c("CDO", "CH4", "FOG", "NOT")
vars_z <- c("GDP")
patterns <- c(vars_y,vars_x,vars_z)


cols_matched <- colnames(df)[
  grepl(paste(patterns, collapse = "|"), colnames(df)) &
  grepl("2019$", colnames(df))
]

df_2019 <- df[, c("code", "iso", "country", cols_matched)]
# How much data are we left with if we remove all rows with NA in any of these columns of interest?

df_2019_clean <- na.omit(df_2019)
dim(df_2019_clean)
summary(df_2019_clean[, cols_matched])
```
It appears that for 2019, which is the last year that we still have fisheries data (FSS), we only have 43 countries that have data for all the variables of interest.

Checking pairwise for countries if we just do a direct X and Y relationship
```{r}
vars_y <- c("PFL", "FSS")
vars_x <- c("CDA", "CHA", "FGA", "NDA")
control <- "GDP"
year <- "2019"

pair_2019 <- do.call(rbind, lapply(vars_y, function(yv) {
  do.call(rbind, lapply(vars_x, function(xv) {
    cols <- c(
      paste0(yv, ".raw.", year),
      paste0(xv, ".raw.", year),
      paste0(control, ".raw.", year)
    )
    
    temp <- df[, c("country", cols)]
    
    data.frame(
      y_var = yv,
      x_var = xv,
      year = as.integer(year),
      n_countries = nrow(na.omit(temp))
    )
  }))
}))

pair_2019[order(-pair_2019$n_countries), ]
```

Checking Year overlaps
```{r}
vars_y <- c("PFL", "FSS")
vars_x <- c("CDA", "CHA", "FGA", "NDA")

get_years <- function(var, data) {
  cols <- colnames(data)[grepl(paste0("^", var, "\\.raw\\."), colnames(data))]
  sub(paste0("^", var, "\\.raw\\.(\\d{4})$"), "\\1", cols)
}

count_pair_nonmissing <- function(y_var, x_var, year, data) {
  y_col <- paste0(y_var, ".raw.", year)
  x_col <- paste0(x_var, ".raw.", year)
  
  if (!(y_col %in% colnames(data)) || !(x_col %in% colnames(data))) return(NA_integer_)
  
  sum(!is.na(data[[y_col]]) & !is.na(data[[x_col]]))
}

pair_results <- do.call(rbind, lapply(vars_y, function(yv) {
  do.call(rbind, lapply(vars_x, function(xv) {
    
    overlap_years <- sort(intersect(get_years(yv, df), get_years(xv, df)))
    
    if (length(overlap_years) == 0) {
      return(data.frame(
        y_var = yv,
        x_var = xv,
        n_overlap_years = 0,
        overlap_years = NA_character_,
        best_year = NA_character_,
        best_n_complete = NA_integer_
      ))
    }
    
    counts <- sapply(overlap_years, function(yr) count_pair_nonmissing(yv, xv, yr, df))
    best_idx <- which.max(counts)
    
    data.frame(
      y_var = yv,
      x_var = xv,
      n_overlap_years = length(overlap_years),
      overlap_years = paste(overlap_years, collapse = ", "),
      best_year = overlap_years[best_idx],
      best_n_complete = counts[best_idx]
    )
  }))
}))

pair_results[order(-pair_results$best_n_complete), ]
```

---

## Variable Selection Decision

```{r variable-selection-stats}
# Compute all stats used in the variable selection narrative below
n_fcl     <- sum(!is.na(df[["FCL.raw.2012"]]))
n_pfl     <- sum(!is.na(df[["PFL.raw.2012"]]))
r_fcl_pfl <- round(cor(df[["FCL.raw.2012"]], df[["PFL.raw.2012"]],
                        use = "complete.obs"), 3)

temp <- data.frame(
  POP     = df[["POP.raw.2012"]],
  CDO = df[["CDO.raw.2012"]],
  NOX = df[["NOX.raw.2012"]],
  SO2 = df[["SO2.raw.2012"]],
  CH4 = df[["CH4.raw.2012"]],
  FCL = df[["FCL.raw.2012"]],
  GDP = df[["GDP.raw.2012"]],
  CDO_pc  = df[["CDO.raw.2012"]] / df[["POP.raw.2012"]],
  NOX_pc  = df[["NOX.raw.2012"]] / df[["POP.raw.2012"]],
  SO2_pc  = df[["SO2.raw.2012"]] / df[["POP.raw.2012"]],
  CH4_pc  = df[["CH4.raw.2012"]] / df[["POP.raw.2012"]],
  FCL_raw = df[["FCL.raw.2012"]],
  GDP_pc  = df[["GPC.raw.2012"]]
)
temp$log_CDO <- log(pmax(temp$CDO,  1e-10))
temp$log_NOX <- log(pmax(temp$NOX,  1e-10))
temp$log_SO2 <- log(pmax(temp$SO2,  1e-10))
temp$log_CH4 <- log(pmax(temp$CH4,  1e-10))
temp$log_FCL <- log(pmax(temp$FCL, 1e-10))
temp$log_GDP <- log(pmax(temp$GDP,  1e-10))
temp$log_POP <- log(pmax(temp$POP,  1e-10))
temp <- na.omit(temp)

r_cdo_fcl <- round(cor(temp$log_CDO, temp$log_FCL), 3)
r_so2_fcl <- round(cor(temp$log_SO2, temp$log_FCL), 3)
r_ch4_fcl <- round(cor(temp$log_CH4, temp$log_FCL), 3)
r_cdo_gdp <- round(cor(temp$log_CDO, temp$log_GDP), 3)
r_cdo_so2 <- round(cor(temp$log_CDO, temp$log_SO2), 3)
r_cdo_ch4 <- round(cor(temp$log_CDO, temp$log_CH4), 3)
r_cdo_nox <- round(cor(temp$log_CDO, temp$log_NOX), 3)

cat("FCL complete cases (2012):", n_fcl, "\n")
cat("PFL complete cases (2012):", n_pfl, "\n")
cat("r(FCL, PFL):", r_fcl_pfl, "\n")
cat("r(log_CDO, log_FCL):", r_cdo_fcl, "\n")
cat("r(log_SO2, log_FCL):", r_so2_fcl, "\n")
cat("r(log_CH4, log_FCL):", r_ch4_fcl, "\n")
cat("r(log_CDO, log_GDP):", r_cdo_gdp, "\n")
cat("r(log_CDO, log_SO2):", r_cdo_so2, "\n")
cat("r(log_CDO, log_CH4):", r_cdo_ch4, "\n")
cat("r(log_CDO, log_NOX):", r_cdo_nox, "\n")
```

**Y: FCL (tree cover loss weighted by permanency)**
FCL gives `r n_fcl` complete cases vs `r n_pfl` for PFL. Both capture
ecosystem degradation from deforestation and are highly correlated
(r = `r r_fcl_pfl`). FCL is preferred purely for statistical power.

**X: CDO per capita (log-transformed)**
CO2 emissions from fossil fuels, normalized by population. We exclude
land use emissions to avoid circularity with our Y variable. Correlation
with log(FCL): r = `r r_cdo_fcl`.

**Robustness X: SO2 per capita (log-transformed)**
Sulfur dioxide emissions from industrial combustion and smelting, a
distinct signal of heavy industrial activity. Correlation with log(FCL):
r = `r r_so2_fcl`. Less collinear with CDO (r = `r r_cdo_so2`) than
GDP, making it a cleaner robustness check.

**Dropped: CH4**
Weakest correlation with log(FCL): r = `r r_ch4_fcl`. High collinearity
with CDO makes it redundant.

**Dropped: GDP as standalone control**
log(CDO per capita) and log(GDP per capita) are correlated at
r = `r r_cdo_gdp`, making GDP nearly redundant as a separate predictor.
Retained only as interaction term to test the EKC hypothesis.

---

## Why CDO Alone Represents Industrial Activity

CDO is highly correlated with all other industrial emissions measures,
meaning it acts as a reliable proxy for the broader industrial footprint
of a country. CDO's strong correlations with NOX (r = `r r_cdo_nox`),
SO2 (r = `r r_cdo_so2`), and GDP (r = `r r_cdo_gdp`) confirm that once
you know a country's CO2 output, you already know a great deal about its
other emissions and wealth.

```{r correlation-heatmap, fig.height=5, fig.width=6}
vars <- c("log_CDO", "log_NOX", "log_SO2",
          "log_CH4", "log_GDP", "log_POP", "log_FCL")

cor_matrix <- cor(temp[, vars], use = "complete.obs")

rownames(cor_matrix) <- c("CDO", "NOX", "SO2", "CH4", "GDP", "POP", "FCL")
colnames(cor_matrix) <- c("CDO", "NOX", "SO2", "CH4", "GDP", "POP", "FCL")

col_palette <- colorRampPalette(c("steelblue", "white", "coral"))(99)

image(
  1:7, 1:7, t(cor_matrix[7:1, ]),
  col = col_palette,
  axes = FALSE,
  xlab = "",
  ylab = "",
  main = "Correlation Matrix — Industrial Activity Variables (2012)"
)

axis(1, at = 1:7, labels = colnames(cor_matrix), las = 2)
axis(2, at = 1:7, labels = rev(rownames(cor_matrix)), las = 1)

for (i in 1:7) {
  for (j in 1:7) {
    text(
      j, 8 - i,
      round(cor_matrix[i, j], 2),
      cex = 0.85,
      font = 2,
      col = ifelse(abs(cor_matrix[i, j]) > 0.6, "white", "black")
    )
  }
}

```

---

## Build Analysis Datasets

Training year: 2012 (138 countries, maximum data). Test year: 2022.

```{r build-datasets}
library(dplyr)
library(stargazer)

build_year_df <- function(data, year) {
  out <- data.frame(
    country = data$country,
    iso     = data$iso,
    POP     = data[[paste0("POP.raw.", year)]],
    CDO     = data[[paste0("CDO.raw.", year)]],
    NOX     = data[[paste0("NOX.raw.", year)]],
    SO2     = data[[paste0("SO2.raw.", year)]],
    CH4     = data[[paste0("CH4.raw.", year)]],
    FCL     = data[[paste0("FCL.raw.", year)]],
    GDP     = data[[paste0("GDP.raw.", year)]],
    CDO_pc  = data[[paste0("CDO.raw.", year)]] / data[[paste0("POP.raw.", year)]],
    NOX_pc  = data[[paste0("NOX.raw.", year)]] / data[[paste0("POP.raw.", year)]],
    SO2_pc  = data[[paste0("SO2.raw.", year)]] / data[[paste0("POP.raw.", year)]],
    CH4_pc  = data[[paste0("CH4.raw.", year)]] / data[[paste0("POP.raw.", year)]],
    FCL_raw = data[[paste0("FCL.raw.", year)]],
    GDP_pc  = data[[paste0("GPC.raw.", year)]],
    HPE     = data[[paste0("HPE.raw.", year)]] # proxy for industry maturity
    #NOD     = data[[paste0("NOD.raw.", year)]] # proxy for industry maturity
    #SOE     = data[[paste0("SOE.raw.", year)]] # proxy for industry maturity
  )
  out$log_CDO <- log(pmax(out$CDO, 1e-10))
  out$log_NOX <- log(pmax(out$NOX, 1e-10))
  out$log_SO2 <- log(pmax(out$SO2, 1e-10))
  out$log_CH4 <- log(pmax(out$CH4, 1e-10))
  out$log_FCL <- log(pmax(out$FCL, 1e-10))
  out$log_GDP <- log(pmax(out$GDP, 1e-10))
  out$log_POP <- log(pmax(out$POP, 1e-10))
  out$log_HPE <- log(pmax(out$HPE, 1e-10)) # proxy for industry maturity
  out$year    <- year
  na.omit(out)
}

df_train <- build_year_df(df, 2012)
df_test  <- build_year_df(df, 2022)

cat("Training set (2012):", nrow(df_train), "countries\n")
cat("Test set    (2022):", nrow(df_test),  "countries\n")
```

---

## Exploratory Data Analysis

### Distributions: Raw vs Log

```{r}

# Turned histogram plotting into a function - are you guys OK with this? If not that's fine.

plot_raw_log_hist <- function(df, var, bins = 25, col = "mediumpurple") {
  
  raw_vals <- df[[var]]
  log_vals <- log(pmax(raw_vals, 1e-10))
  
  par(mfrow = c(1, 2))
  
  # Raw
  hist(raw_vals,
       breaks = bins,
       col = col,
       border = NA,
       main = paste0(var, " — Raw | mean = ",
                     round(mean(raw_vals, na.rm = TRUE), 2)),
       xlab = var)
  
  abline(v = mean(raw_vals, na.rm = TRUE),
         col = "red", lwd = 2, lty = 2)
  
  # Log
  hist(log_vals,
       breaks = bins,
       col = col,
       border = NA,
       main = paste0(var, " — Log | mean = ",
                     round(mean(log_vals, na.rm = TRUE), 2)),
       xlab = paste0("log(", var, ")"))
  
  abline(v = mean(log_vals, na.rm = TRUE),
         col = "red", lwd = 2, lty = 2)
  
  par(mfrow = c(1, 1))
}

plot_raw_log_hist(df_train, "FCL_raw")
plot_raw_log_hist(df_train, "CDO")
plot_raw_log_hist(df_train, "POP")
plot_raw_log_hist(df_train, "GDP")
plot_raw_log_hist(df_train, "SO2")
plot_raw_log_hist(df_train, "HPE")


```


```{r eda-distributions-fcl, fig.height=3.5, fig.width=9}
par(mfrow = c(1, 2))

hist(df_train$FCL_raw, breaks = 25, col = "mediumpurple", border = NA,
     main = "FCL — Raw", xlab = "Tree Cover Loss (proportion)")
abline(v = mean(df_train$FCL_raw), col = "red", lwd = 2, lty = 2)

hist(df_train$log_FCL, breaks = 25, col = "mediumpurple", border = NA,
     main = paste0("FCL — Log  |  mean = ",
                   round(mean(df_train$log_FCL), 2)),
     xlab = "log(Tree Cover Loss)")
abline(v = mean(df_train$log_FCL), col = "red", lwd = 2, lty = 2)

par(mfrow = c(1, 1))
```

```{r eda-distributions-cdo, fig.height=3.5, fig.width=9}
par(mfrow = c(1, 2))

hist(df_train$CDO, breaks = 25, col = "steelblue", border = NA,
     main = "CO2 — Raw", xlab = "CO2 (Gg)")
abline(v = mean(df_train$CDO), col = "red", lwd = 2, lty = 2)

hist(df_train$log_CDO, breaks = 25, col = "steelblue", border = NA,
     main = paste0("CO2 per Capita — Log  |  mean = ",
                   round(mean(df_train$log_CDO), 2)),
     xlab = "log(CO2 per capita)")
abline(v = mean(df_train$log_CDO), col = "red", lwd = 2, lty = 2)

par(mfrow = c(1, 1))
```

```{r eda-distributions-so2, fig.height=3.5, fig.width=9}
par(mfrow = c(1, 2))

hist(df_train$SO2_pc, breaks = 25, col = "coral", border = NA,
     main = "SO2 per Capita — Raw", xlab = "SO2 per capita (Gg/person)")
abline(v = mean(df_train$SO2_pc), col = "red", lwd = 2, lty = 2)

hist(df_train$log_SO2, breaks = 25, col = "coral", border = NA,
     main = paste0("SO2 per Capita — Log  |  mean = ",
                   round(mean(df_train$log_SO2), 2)),
     xlab = "log(SO2 per capita)")
abline(v = mean(df_train$log_SO2), col = "red", lwd = 2, lty = 2)

par(mfrow = c(1, 1))
```

```{r eda-distributions-gdp, fig.height=3.5, fig.width=9}
par(mfrow = c(1, 2))

hist(df_train$GDP_pc, breaks = 25, col = "seagreen", border = NA,
     main = "GDP per Capita — Raw", xlab = "GDP per capita (USD)")
abline(v = mean(df_train$GDP_pc), col = "red", lwd = 2, lty = 2)

hist(df_train$log_GDP, breaks = 25, col = "seagreen", border = NA,
     main = paste0("GDP per Capita — Log  |  mean = ",
                   round(mean(df_train$log_GDP), 2)),
     xlab = "log(GDP per capita)")
abline(v = mean(df_train$log_GDP), col = "red", lwd = 2, lty = 2)

par(mfrow = c(1, 1))
```

### Scatterplots: X vs Y

```{r eda-scatter, fig.height=4, fig.width=11}
par(mfrow = c(1, 3))

plot(df_train$log_CDO, df_train$log_FCL,
     xlab = "log(CO2 per capita)", ylab = "log(Tree Cover Loss)",
     main = paste0("CO2 vs FCL  |  r = ",
                   round(cor(df_train$log_CDO, df_train$log_FCL), 3)),
     pch = 19, col = rgb(0.2, 0.4, 0.8, 0.5), cex = 0.8)
abline(lm(log_FCL ~ log_CDO, data = df_train), col = "red", lwd = 2)

plot(df_train$log_SO2, df_train$log_FCL,
     xlab = "log(SO2 per capita)", ylab = "log(Tree Cover Loss)",
     main = paste0("SO2 vs FCL  |  r = ",
                   round(cor(df_train$log_SO2, df_train$log_FCL), 3)),
     pch = 19, col = rgb(0.8, 0.4, 0.2, 0.5), cex = 0.8)
abline(lm(log_FCL ~ log_SO2, data = df_train), col = "red", lwd = 2)

plot(df_train$log_GDP, df_train$log_FCL,
     xlab = "log(GDP per capita)", ylab = "log(Tree Cover Loss)",
     main = paste0("GDP vs FCL  |  r = ",
                   round(cor(df_train$log_GDP, df_train$log_FCL), 3)),
     pch = 19, col = rgb(0.2, 0.7, 0.4, 0.5), cex = 0.8)
abline(lm(log_FCL ~ log_GDP, data = df_train), col = "red", lwd = 2)

par(mfrow = c(1, 1))
```

### Correlation Matrix (Training Set)

```{r eda-correlation}
cor_vars <- df_train[, c("log_CDO", "log_SO2", "log_GDP", "log_POP", "HPE", "log_FCL")]
round(cor(cor_vars, use = "complete.obs"), 3)
pairs(cor_vars)
```

---

## Models

### Model 1: Simplest credible model (CDO + POP)

```{r model1}

m1 <- lm(log_FCL ~ log_CDO + log_POP, data = df_train)
summary(m1)
```

### Model 2: CDO + GDP + POP

```{r model2}
m2 <- lm(log_FCL ~ log_CDO + log_GDP + log_POP, data = df_train)
summary(m2)
```

### Model 2b: using HPE instead of GDP

### Potentially HPE (human pollution exposure) can be an indicator of 'first-world' country behavior

```{r model2}

m2b <- lm(log_FCL ~ log_CDO + log_POP + HPE, data = df_train)
summary(m2b)

plot(m2b)
print(round(vif(m2b), 3))
print(bptest(m2b))
```

### Added SO2 but this is very correlated to CDO, probably not ideal

```{r}

m2c <- lm(log_FCL ~ log_CDO + log_POP + log_SO2 + HPE, data = df_train)
summary(m2c)
print(round(vif(m2b), 3))
print(bptest(m2b))
plot(m2c)

```

### Model 3b: SO2 as primary X (replaces CDO)

```{r model3b}
m3b <- lm(log_FCL ~ log_SO2 * log_GDP, data = df_train)
summary(m3b)
```

### Regression Table

```{r regression-table, results='asis'}
stargazer(m1, m2, m3a, m3b,
          type             = "text",
          title            = "Tree Cover Loss on Industrial Activity (Training Year: 2012)",
          dep.var.labels   = "log(Tree Cover Loss)",
          covariate.labels = c("log(CO2 per capita)",
                               "log(GDP per capita)",
                               "log(SO2 per capita)",
                               "log(CO2) x log(GDP)",
                               "log(SO2) x log(GDP)"),
          omit.stat        = c("f", "ser"),
          star.cutoffs     = c(0.05, 0.01, 0.001))
```

---

## Model Diagnostics (Model 2)

```{r diagnostics, fig.height=7, fig.width=10}
par(mfrow = c(2, 2))
plot(m2, which = 1, main = "Residuals vs Fitted")
plot(m2, which = 2, main = "Q-Q Plot")
plot(m2, which = 3, main = "Scale-Location")
plot(m2, which = 5, main = "Residuals vs Leverage")
par(mfrow = c(1, 1))
```

```{r assumption-tests}
cat("=== VIF (Model 2) ===\n")
print(round(vif(m2), 3))

cat("\n=== Breusch-Pagan Test (Model 2) ===\n")
print(bptest(m2))
```

---

## Out-of-Sample Test (2022)

```{r out-of-sample, fig.height=4, fig.width=6}
df_test$predicted <- predict(m2b, newdata = df_test)

rmse_train <- round(sqrt(mean(residuals(m2)^2)), 4)
rmse_test  <- round(sqrt(mean((df_test$log_FCL - df_test$predicted)^2,
                               na.rm = TRUE)), 4)

ss_res <- sum((df_test$log_FCL - df_test$predicted)^2, na.rm = TRUE)
ss_tot <- sum((df_test$log_FCL - mean(df_test$log_FCL))^2, na.rm = TRUE)
r2_test <- round(1 - ss_res / ss_tot, 4)

cat("RMSE (train, 2012):", rmse_train, "\n")
cat("RMSE (test,  2022):", rmse_test,  "\n")
cat("R2   (train)      :", round(summary(m2)$r.squared, 4), "\n")
cat("R2   (test,  2022):", r2_test, "\n")

plot(df_test$predicted, df_test$log_FCL,
     xlab = "Predicted log(FCL)", ylab = "Actual log(FCL)",
     main = paste0("Out-of-Sample: 2022\n",
                   "RMSE train = ", rmse_train,
                   "  |  RMSE test = ", rmse_test),
     pch = 19, col = rgb(0.2, 0.6, 0.4, 0.6))
abline(0, 1, col = "red", lwd = 2, lty = 2)
```