---
  title: "mids-203-project"
output: html_document
date: "2026-04-09"
---
  ```{r cars}
# Load libraries
library(ggplot2)

# Read CSV (adjust path as needed)
df <- read.csv("merged_raw_all_years.csv")
dim(df)
```

First task: examine the data for cleanliness issues

- Realizing that a lot of the variables in the dataset are missing like 50% of the rows...so let's first rank all the possible Y variables (in 'ecosystem vitality on page 3') and X variables (in 'climate change on page 4') based on how many NA rows there are, then choose based on a tradeoff between relevance and # of missing rows.
- I'm using 2019, which is the last year with fisheries data (see page 130/131)

```{r}
ecosystem_vars <- c(
  # Biodiversity & Habitat
  "MKP", "MHP", "MPE", "PAR", "SPI", "TBN", "TKP", "PAE", "PHL", "RLI", "SHI", "BER",
  
  # Forests
  "PFL", "IFL", "FCL", "TCG", "FLI",
  
  # Fisheries
  "FSS", "FCD", "BTZ", "BTO", "RMS",
  
  # Air Pollution
  "OEB", "OEC", "NXA", "SDA",
  
  # Agriculture
  "SNM", "PSU", "PRS", "RCY",
  
  # Water Resources
  "WWG", "WWC", "WWT", "WWR"
)

eco_col_matches <- colnames(df)[
  grepl(paste(ecosystem_vars, collapse = "|"), colnames(df)) &
    grepl("2019$", colnames(df))
]

df_ecosystem = df[, eco_col_matches]

# sort each column in the ecosystem vitality (ECO) category by the number of NAs in the year 2019
sort(colSums(is.na(df_ecosystem)))

```
Most relevant Y-variables would be:
  - PFL (primary forest loss) - 137 missing rows out of 220
- Alternative: FCL (74 missing rows)
- FSS (fish stock status) - 102 missing rows out of 220
- Alternative: RLI/SHI - 2 and 43 missing rows
- Could consider using TBN (terrestrial biome protection/biodiversity/ecosystem vitality) - not ideal but it has no missing data

Now to look at the x-variables:
  
  ```{r}
climate_vars <- c(
  # Climate Change Mitigation
  "CDA",  # Adjusted emissions growth rate for CO2
  "CDF",  # CO2 growth rate (country-specific targets)
  "CHA",  # Adjusted emissions growth rate for methane
  "FGA",  # Adjusted emissions growth rate for F-gases
  "NDA",  # Adjusted emissions growth rate for nitrous oxide
  "BCA",  # Adjusted emissions growth rate for black carbon
  "LUF",  # Net carbon fluxes due to land cover change
  "GTI",  # GHG growth rate adjusted by emissions intensity
  "GTP",  # GHG growth rate adjusted by per capita emissions
  "GHN",  # Projected emissions in 2050
  "CBP",   # Projected cumulative emissions to 2050 relative to carbon budget,
)

climate_col_matches <- colnames(df)[
  grepl(paste(climate_vars, collapse = "|"), colnames(df)) &
    grepl("2019$", colnames(df))
]

df_climate = df[, climate_col_matches]

# sort each column in the ecosystem vitality (ECO) category by the number of NAs in the year 2019
sort(colSums(is.na(df_climate)))


```

The x-variables we might use are basically all missing 20-30 rows, not bad. We can then choose based purely on what's most relevant:
- CDO (carbon dioxide emissions)
- CH4 (methane emissions)
- FOG (F-gas emissions)
- NOT (N2O emissions)

We can also investigate how much missing data is in the GDP for 2019:
```{r}
sum(is.na(df[, 'GDP.raw.2019']))
```

What if we looked at only our best variables of interest and saw how small that dataset would be?

```{r}
vars_y <- c("PFL", "FSS")
vars_x <- c("CDO", "CH4", "FOG", "NOT")
vars_z <- c("GDP")
patterns <- c(vars_y,vars_x,vars_z)


cols_matched <- colnames(df)[
  grepl(paste(patterns, collapse = "|"), colnames(df)) &
  grepl("2019$", colnames(df))
]

df_2019 <- df[, c("code", "iso", "country", cols_matched)]
# How much data are we left with if we remove all rows with NA in any of these columns of interest?

df_2019_clean <- na.omit(df_2019)
dim(df_2019_clean)
summary(df_2019_clean[, cols_matched])
```
It appears that for 2019, which is the last year that we still have fisheries data (FSS), we only have 43 countries that have data for all the variables of interest.

